---
title: 'Replication Code for Figures in `A Practical Guide to Dealing with Attrition in Political Science Experiments`'
date: 'This version: February 2022'
output: 
  html_document: 
    pandoc_args: ["--lua-filter=color-text.lua"]
    toc: true
    toc_float: true
    number_sections: true
    theme: united
  pdf_document: 
    pandoc_args: ["--lua-filter=color-text.lua"]
    keep_tex: true
header-includes:
  - \preauthor{\centering\large}
  - \predate{\centering\normalsize}
  - \pretitle{\centering\Large\textbf}
  - \usepackage{setspace}\onehalfspacing
  - \usepackage{xcolor}
  - \usepackage[shortlabels]{enumitem}
  - \usepackage{pgf,tikz, mathabx}
  - \usetikzlibrary{positioning}
editor_options:
  chunk_output_type: console
---
<style>

table, td, th {
  border: none;
  padding-left: 1em;
  padding-right: 1em;
  margin-left: auto;
  margin-right: auto;
  margin-top: 1em;
  margin-bottom: 1em;
}

</style>

```{cat, engine.opts = list(file = "color-text.lua")}
Span = function(span)
  color = span.attributes['color']
  -- if no color attribute, return unchange
  if color == nil then return span end
  
  -- tranform to <span style="color: red;"></span>
  if FORMAT:match 'html' then
    -- remove color attributes
    span.attributes['color'] = nil
    -- use style attribute instead
    span.attributes['style'] = 'color: ' .. color .. ';'
    -- return full span element
    return span
  elseif FORMAT:match 'latex' then
    -- remove color attributes
    span.attributes['color'] = nil
    -- encapsulate in latex code
    table.insert(
      span.content, 1,
      pandoc.RawInline('latex', '\\textcolor{'..color..'}{')
    )
    table.insert(
      span.content,
      pandoc.RawInline('latex', '}')
    )
    -- returns only span content
    return span.content
  else
    -- for other format return unchanged
    return span
  end
end
```

<!-- # Packages: -->

```{r,echo=FALSE,eval=TRUE,message=FALSE,warning=FALSE}
rm(list=ls())
library(viridis)
require(tidyverse)
require(waffle)
require(grid)
require(ggpattern)
require(ggpubr)

source("../attritevis_fns.R")
```

Code to replicate figures from the paper A Practical Guide to Dealing with Attrition in Political Science Experiments by Adeline Lo, Jonathan Renshon, and Lotem Bassan-Nygate.


# Fiure 1: Experimental papers in fullJEPScorpus and their discussion of attrition

```{r waffle, echo=TRUE, eval=TRUE, message=FALSE,warning=FALSE}
#Reading in CSV Data
attrition <- read_csv("lit_review.csv")


#Functions to remove "*" and change "Yes" to 1 and "No" to 0
remove_star <- function(x) {
  return(str_extract(x, "Yes|No"))
}

yesno_onezero <- function(x) {
  return(case_when(x == "Yes" ~ 1,
                   x == "No" ~ 0))
}

attrition <- attrition %>% 
  mutate_at(c(7:14), remove_star) %>%
  mutate_at(c(7:14), yesno_onezero)


#Creating table of proportions
prop_att <- mean(attrition$Attrition)
prop_noatt <- mean(attrition$`0 Attrition`[attrition$Attrition == 1])
prop_attdv <- mean(attrition$`Response Rate DV`[attrition$Attrition == 1])
prop_quan <- mean(attrition$`Quantified Attrition`[attrition$Attrition == 1 & attrition$`0 Attrition` == 0 & attrition$`Response Rate DV` == 0])
prop_adj <- mean(attrition$`Sample Adjustments`[attrition$Attrition == 1 & attrition$`0 Attrition` == 0 & attrition$`Response Rate DV` == 0])

attrition_summary <- as_tibble(data.frame(
  c("Measurement",
    "Proportion that mention attrition",
    "Proportion \"no attrition\"",
    "Proportion DV",
    "Proportion quantify",
    "Proportion adjust"),
  c("Value",
    prop_att,
    prop_noatt,
    prop_attdv,
    prop_quan,
    prop_adj)
))


#Creating variable for the waffle plot
count <- attrition %>% 
  mutate(waffle = case_when(`Sample Adjustments` == 1 ~ "Attrition mentioned, quantified, analyzed",
                            Attrition == 1 & `0 Attrition` == 0 & `Response Rate DV` == 0 & `Sample Adjustments` == 0 & `Quantified Attrition` == 1 ~ "Attrition mentioned and quantified",
                            Attrition == 1 & `0 Attrition` == 0 & `Response Rate DV` == 0 & `Sample Adjustments` == 0 & `Quantified Attrition` == 0 ~ "Attrition mentioned only",
                            `Response Rate DV` == 1 ~ "Attrition is DV",
                            `0 Attrition` == 1 ~ "Attrition mentioned - none in study",
                            Attrition == 0 ~ "No mention of attrition")) %>% 
  group_by(waffle) %>%
  summarise(n = n())

#Reordering to make legend easier to read and plot look better
count <- count[c(5,3,2,4,1,6),]

#Creating waffle plot
case_counts <- count$n
names(case_counts) <- count$waffle

waffle(case_counts, colors = c(
  "#fcba03", #For Attrition mentioned, quantified, analyzed
  "#e8803f", #For Attrition mentioned and quantified
  "#965ef7", #For Attrition mentioned, none in study
  "#595959", #For Attrition mentioned only
  "#5eccf7", #For Attrition is DV
  "#ff6666"  #For No mention of attrition
  )) +
  theme(legend.key.size = unit(10, "mm"), legend.text = element_text(size = 12))

```

# Figure 3: Attrition timeline visualization

```{r timeline, echo=TRUE, eval=TRUE, message=FALSE,warning=FALSE}

plot_withbars <- function(data, title)
{ 
  #required packages
  require(ggplot2)
  require(viridis)
  require(Hmisc)
  require(dplyr)
 
  #Begin by creating an attrition dataframe
  #for each missing value in the dataframe `dataset` give value 1, otherwise give 0.
  data <- apply(data,2,function(x) {ifelse(is.na(x),1,0)})
  
  #change `skippers` into 0 (we are only interested in respondents that attrited).
  
  data<-apply(data,1,skip_to_attrite)
  data<-t(data) #transpose data
  
  data2<-data.frame(data)
  
  #transform into a long dataframe, such that the variable `attrited` is the number of missing observations per variable.
  data <- data.frame(colSums(data2))
  colnames(data) <- "attrited"
  
  #transform `attrited` to measure how many respondents attrited during each question, rather than how many missing values are in each question.
  attrite_2<-data$attrited
  num_dropped <- data[-1,] - data[-nrow(data),]
  data$attrited<- c(data[1,], num_dropped)
  data$attrite_2<-attrite_2
  
  #add variable `proportion` = number of attrited respondents / starting N.
  data$n_prev <- NULL
  data$attrite_2 <- NULL
  data$proportion <-    round(data$attrited/nrow(data2),2)
  
  #add variable `questions` = the name of each variable in the original dataframe.
  data$questions <- rownames(data)
  rownames(data) <- c()
  
  
    tmp_colors<-viridis(n=2,alpha=0.6,begin=0.25,end=1,direction=1,option="D")
    data$questions <- factor(data$questions, levels=data$questions)
    data$questions<-as.numeric(data$questions)

p <- data %>%
  
  ggplot(aes(questions,proportion)) +
  
  
  geom_rect_pattern(aes(xmin= 4.7,
                        xmax = 10,
                        ymin = 0,
                        ymax = Inf), 
                    pattern = 'stripe',
                    pattern_fill = 'pink',
                    pattern_colour ='pink',
                    pattern_size = 3.5,
                    pattern_alpha = 0.02,
                    pattern_density = 0.02,
                    alpha = 0.02,
                    fill    = 'white') +
  
  geom_rect_pattern(aes(xmin=1,
                        xmax = 4.5,
                        ymin = 0,
                        ymax = Inf), 
                    pattern = 'stripe',
                    pattern_fill = 'gray',
                    pattern_colour ='gray',
                    pattern_size = 3.5,
                    pattern_alpha = 0.02,
                    pattern_density = 0.02,
                    alpha = 0.02,
                    fill    = 'white') +
  
  #add if statement based on freq
  
  geom_histogram(color="#e9ecef", alpha=0.6, stat = 'identity') +
  
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10),
                     labels=c("Q1", "Q2", "Q3", "Q4", "Q5","Q6","Q7","Q8"
                              ,"Q9","Q10"))

scale_fill_manual(values=tmp_colors) 

#vlines

pre_treatment <- as.numeric(c(1,2,3,4))
treatment <- as.numeric(5)
DV<- as.numeric(c(6,7))
post_treatment <- as.numeric(c(8,9,10))

p <- p + geom_vline(data= data.frame(type="Pre-Treatment", 
                                     col="Pre-Treatment", pre_treatment = pre_treatment),
                    aes(colour=col, xintercept = pre_treatment), 
                    size = 0.7, show.legend = TRUE)

p <- p + geom_vline(data= data.frame(type="Treatment", 
                                     col="Treatment", treatment = treatment),
                    aes(colour=col, xintercept = treatment), 
                    size = 2, show.legend = TRUE)

p <- p + geom_vline(data= data.frame(type="Outcome", 
                                     col="Outcome", DV = DV),
                    aes(colour=col, xintercept = DV), 
                    size = 0.7, show.legend = TRUE)

p <- p + geom_vline(data= data.frame(type="Post-Treatment", 
                                     col="Post-Treatment", post_treatment = post_treatment),
                    aes(colour=col, xintercept = post_treatment), 
                    size = 0.7, show.legend = TRUE)


p <- p + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
               panel.background = element_blank(),
               axis.text.x = element_text(angle = 90, hjust = 1, size = 8))   

#add legend details manually  
p<- p + scale_colour_manual(name="Legend"
                            ,breaks = c("Pre-Treatment","Treatment",
                                        "Outcome","Post-Treatment")
                            ,labels = c("Pre-Treatment","Treatment",
                                        "Outcome","Post-Treatment")
                            ,values = c("goldenrod3","firebrick","royalblue3","seagreen")
) +
  labs(x = "Survey Questions") + #titles
  labs(y = "Proportion of respondents attrited") +
  ggtitle(title) + ylim(0,1) 

}


#Attition post-treatment (throughout survey)
n <- 1000
df <- data.frame(
  Q1 = sample(c(18:90), n, rep = TRUE), #age
  Q2 = sample(c("m", "f"), n, rep = TRUE, prob = c(0.55, 0.45)), #sex
  Q3 = sample(c(0,1), n, rep = TRUE), #other general pre-treatment questions
  Q4 = sample(c(0,1), n, rep = TRUE),
  Q5 = sample(c("treatment", "control"), n, rep = TRUE), #we will assume Q5 is the treatment
  Q6 = sample(c(0,1), n, rep = TRUE), #post treatment questions
  Q7 = sample(c(0,1), n, rep = TRUE),
  Q8 = sample(c(0,1), n, rep = TRUE),
  Q9 = sample(c(0,1), n, rep = TRUE),
  Q10 = sample(c(0,1), n, rep = TRUE))

#repeat for 4 plots
df1<-df
df2<-df
df3<-df
df4<-df

#Generate attrition post
invisible(
  sapply(sample(1:nrow(df1), 150),function(x) {
    a <- sample(1:10,1)
    df1[x,a:ncol(df1)] <<- NA
  }
  ))

a<-plot_withbars(data=df1, 
                 title = "(a) Low Level Attrition")


#Generate attrition pre-treatment
invisible(
  sapply(sample(1:nrow(df2), 450),function(x) {
    a <- sample(1:4,1)
    df2[x,a:ncol(df2)] <<- NA
  }
  ))

b<-plot_withbars(data=df2, 
                 title ="(b) Pre-treatment Attrition")



#Generate post-treatment attrition (immediate)

invisible(
  sapply(sample(1:nrow(df3), 450),function(x) {
    a <- sample(5:6,1)
    df3[x,a:ncol(df3)] <<- NA
  }
  ))

c<-plot_withbars(data=df3,
                 title ="(c) Post-treatment Attrition (immediate)")


#Generate post-treatment attrition (prolonged)
invisible(
  sapply(sample(1:nrow(df4), 450),function(x) {
    a <- sample(6:10,1)
    df4[x,a:ncol(df4)] <<- NA
  }
  ))

d<-plot_withbars(data=df4,
                 title ="(d) Post-treatment Attrition (prolonged)")


#combine plots to one figure
figure <- ggarrange(a + rremove("ylab") + rremove("xlab"), b + rremove("ylab") 
                    + rremove("xlab"), c + rremove("ylab") + rremove("xlab"), 
                    d + rremove("ylab") + rremove("xlab"), # remove axis labels from plots
                    labels = NULL,
                    ncol = 2, nrow = 2,
                    common.legend = TRUE, legend = "top",
                    align = "hv", 
                    font.label = list(size = 10, color = "black", face = "bold", 
                                      family = NULL, position = "top"))

annotate_figure(figure, left = textGrob("Proportion of attrited", rot = 90, vjust = 1, gp = gpar(cex = 1.5)),
                bottom = textGrob("Experiment Questions", gp = gpar(cex = 1.5)))

```
